---
title: 'Generalized Ideal Point Models for Noisy Dynamic Measures in the Social Sciences'
author: 
  - name: "Robert Kubinec"
    affiliations: 
        - ref: usc
affiliations:
    - id: usc
      name: "University of South Carolina"
date: last-modified
execute: 
  cache: true
  echo: false
  warning: false
  error: false
date-format: long
format:
  wordcount-pdf: 
    wordcount-banner: true
    header-includes: |
      \usepackage[margin=1in]{geometry}
      \usepackage{setspace}
    fig-height: 3.5
    fig-width: 6.5
filters:
  - authors-block
abstract: "This paper presents novel specifications for non-ignorable missingness and flexible time-series specifications  within a robust Bayesian framework. To do so, the paper develops computational techniques for massive parallel computation of dynamic ideal points using Stan, a state-of-the-art estimation engine for Bayesian models. The proposed specification is compared to existing dynamic ideal point models using Monte Carlo simulations to show that the model is able to minimize specification error arising from selection bias and time-series autocorrelation. To show the model's utility, I estimate monthly trajectories from 1990 to 2018 for all U.S. House Congresspeople, which represents a twelve-fold increase in the granularity of ideal point measurements relative to existing methods and is also robust to strategic abstention from votes. WORD COUNT: {{< words-sum body-note >}}"
bibliography: "BibTexDatabase.bib"
fig-cap-location: top
csl: apsr.csl
---

```{r setup, include=FALSE}

require(dplyr)
require(ggplot2)
require(tidyr)
require(readr)
require(lubridate)
require(idealstan)
require(forcats)
library(patchwork)
library(posterior)
library(marginaleffects)
library(tinytable)

# set to true to create all analyses from fitted idealstan models
# requires significant time and RAM (especially the latter)
run_all <- F

# see file to_cluster.R for the fitting of the actual idealstan models

# load the datas

rollcalls <- readRDS('data/rollcalls.rds')

unemp1 <- rollcalls %>% 
  select(cast_code,rollnumber,congress,
         bioname,party_code,date_month,unemp_rate) %>% 
  mutate(item=paste0(congress,"_",rollnumber),
         cast_code=recode_factor(cast_code,Abstention=NA_character_),
         cast_code=as.numeric(cast_code)-1,
         bioname=factor(bioname),
         unemp_rate=100*unemp_rate,
         bioname=relevel(bioname,"DeFAZIO, Peter Anthony")) %>% 
  distinct %>% 
  filter(party_code %in% c("R","D")) %>% 
  mutate(party_code=factor(party_code))

# drop legislators who vote on fewer than 25 unanimous bills

check_bills <- group_by(unemp1,item,party_code,cast_code) %>% count %>% 
  group_by(item,party_code) %>% 
  summarize(prop=n[cast_code==1] / sum(n),
            n_vote=sum(n)) %>% 
  ungroup %>% 
  mutate(prop=ifelse(prop==.5,
                          sample(c(.49,.51),1),
                     prop),
    util_func=(1 / sqrt((.5 - prop)^2))*n_vote) %>% 
  arrange(desc(util_func))

legis_count <- group_by(unemp1, item) %>% 
  mutate(unan=all(cast_code[!is.na(cast_code)]==1) || all(cast_code[!is.na(cast_code)]==0)) %>% 
  group_by(bioname) %>% 
  summarize(n_votes_nonunam=length(unique(item[!unan])))

# check number of days in legislature

num_days <- distinct(unemp1,bioname,date_month) %>% 
  count(bioname)

```

\newpage

\doublespacing

This paper shows why measurement models need to become more robust and flexible in order to handle the difficult inference challenges of contemporary social science scholarship with noisy dynamic data. To accomplish this task, I define a generalization of the ideal point model that employs novel Bayesian estimation techniques [@CarpenterGelmanHoffmanEtAl2017] for massive and robust computation. The goal of this modeling framework is to more closely match latent variable specifications with theoretical priors in order to reduce mis-specification challenges related to missing data and time series processes. In addition, the `idealstan` R package implements these new methods with advances in Bayesian computation to handle datasets that are several orders of magnitude larger than was previously feasible with robust Bayesian uncertainty in latent scores.

Initially designed for measuring U.S. Congressional ideology, the ideal point model has proven valuable for various social science measurement challenges, including Twitter data [@barbera2015; @eadyNewsSharingSocial2025], state capacity [@hansonLeviathansLatentDimensions2021], judiciaries [@ovádek2021], lobbying [@kim2017], the level of democracy [@pemsteinDemocraticCompromiseLatent2017; @grumbachLaboratoriesDemocracyHow2022], and large language model output [@burnham], to name only a few. Its utility lies in measuring latent traits where data are imperfect proxies of theoretically defined concepts. However, while social scientists increasingly have access to granular time-varying data, dynamic ideal point modeling is limited to measures of binary outcomes, very specific time-series processes, fully observed data, and depending on the estimation framework, datasets of small to medium size. As a result, scholars often must choose a measurement model with assumptions that are inappropriate for the measurement problem at hand, which can perversely lead to new measurement error via model mis-specification. On the other hand, if scholars choose to ignore measurement estimation altogether, they can end up with fragile inferences that depend on arbitrary measurement artifacts [@jerzak2024].

<!-- The new measurement framework proposed in this paper offers scholars latent variable estimation that avoids these problems by advancing new methods for complex dynamic processes, non-ignorable missingness, and large datasets. The main advantage of this framework is that it enables researchers to overcome vexing measurement challenges and avoid downstream inferential errors.  -->

To show how `idealstan` advances our understanding of measurement models, I engage in both simulations and empirical innovation. I first analyze Monte Carlo simulations of the `idealstan` method compared to existing approaches. As I show, `idealstan` is notably faster than existing Bayesian implementations but also avoids consequential inferential errors due to a lack of comparable time series processes in existing work. In addition, I show that non-ignorable missingness, specifically a form of selection bias, can have profound consequences for latent variable estimation in dynamic settings.

Second, I show how `idealstan` can identify much more granular and theoretically-precise latent traits than previous models. To that end, I estimate *monthly* ideal points for *all* legislators in the U.S. House from 1990 to 2018, which represents a twelve-fold increase in granularity in ideal point measurements relative to existing measures. As I show, increasing measurement granularity requires selecting an optimal time-series specification in order to identify a specification that most closely matches the theoretical nature of the construct of interest. The empirical reveals that strategic abstention tends to occur more frequently at the end of congressional sessions, suggesting that ideal points are not as static within Congressional sessions as existing models show.

By combining missing data adjustments, time-varying inference, and scalable computation, `idealstan` provides a robust method for addressing next-generation dynamic measurement challenges across the social sciences.

# Statistical Measurement in the Social Sciences

In the social sciences, measurement error is almost always non-trivial. For example, voting records, which are one of the most directly observable forms of political data, are nonetheless riddled with inaccurate records and duplicates [@kimEvaluatingQualityChanges2019]. Measurement error is more commonly recognized with higher-level concepts like democracy that have led to ongoing scholarly debates [@littleMeasuringDemocraticBacksliding2024; @knutsenConceptualMeasurementIssues2024]. However, virtually all social science variables have at least some measurement error because collecting data on humans, who have agency that allows them to hide or misreport information, is never a trivial matter. While many scholars recognize these structural limitations, and measurement models have a long history in statistics [@flake2019], it can be relatively difficult to apply measurement models to real-world datasets, especially when the tool needs to be able to incorporate relevant theoretical priors about the latent trait.

Building a theoretically-valid measurement model can be a very useful contribution to our ability to analyze concepts, such as the influential work on legislative ideology in the United States [@poole1997], but deriving an appropriate model can involve extensive technical work to incorporate theoretical priors [@caugheyDynamicEstimationLatent2015]--if it is even possible at all. The model I present in this paper, which I shall refer to as `idealstan` as that is the name of the associated software package, aims to contribute a general purpose measurement model that offers more flexibility and robustness in capturing latent concepts. I argue that this type of flexibility is very important for latent variables because model assumptions about the nature of latent traits can have important consequences for downstream inference. Ideally, a parametric measurement model should fit as close as possible to researcher priors while incorporating as much data as possible to reduce measurement error. Given the increasingly widespread availability of time-varying data in the social sciences, the focus in this paper is on the unique challenges that arise with *dynamic* measurement modeling—an area in which there are relatively few options for scholars to work with.

<!-- In the social sciences, measurement error is almost always non-trivial. For example, voting records, which are one of the most directly observable forms of political data, are nonetheless riddled with inaccurate records and duplicates [@ansolabehereQualityVoterRegistration2010]. Measurement error is more commonly recognized with higher-level concepts like democracy that have led to ongoing scholarly debates [@littleMeasuringDemocraticBacksliding2024; @millerHowLittleMengs2024; @knutsenConceptualMeasurementIssues2024]. However, virtually all social science variables have at least some measurement error because collecting data on humans, who have agency that allows them to hide or misreport information, is never a trivial matter. While many scholars recognize these structural limitations, and measurement models have a long history in statistics [@flake2019], it can be relatively difficult to apply measurement models to real-world datasets, especially when the tool needs to be able to incorporate relevant theoretical priors about the latent trait. Building a theoretically-valid measurement model can be a very useful contribution to our ability to analyze concepts, such as the influential work on legislative ideology in the United States [@poole1997; @jackman2004], but deriving an appropriate model can involve extensive technical work to incorporate theoretical priors [@caugheyDynamicEstimationLatent2015]. The model I present in this paper, which I shall refer to as `idealstan` as that is the name of the associated software package, aims to contribute a general purpose measurement model that applies to diverse datasets and problems while also permitting the incorporation of theory into measurement choices. By doing so, I hope to mainstream relatively complex Bayesian measurement models that have often taken months or years to derive and permit them to be rapidly deployed to diverse phenomena. -->

The specific parameterization adopted in this paper is based on the ideal point model because I argue that ideal point models contain a nuanced balance of exploratory (or unsupervised) and confirmatory (or supervised) aspects. As a result, ideal point models can be repurposed to a wide variety of measurement domains where researchers have at least some theoretical priors about the latent concept of interest. Ideal point models have a long history in political science, and time-varying ideal point models have been available for almost as long. Latent measurements derived from these models, including DW-NOMINATE scores for the U.S. Congress [@poole1997] and comparable measures for the U.N. General Assembly, democracy, state capacity, and human rights compliance [@mesquita2022; @gauthier2025; @hansonLeviathansLatentDimensions2021; @ari2025; @gaughan2024], have made an outsize impact on research in the discipline. These uni-dimensional estimates of hard-to-measure traits permit researchers to test hypotheses with over-time variation to explore the factors behind evolutionary changes. While over-time variation is no substitute for a causally-identified experiment, it can help rule out important inference issues like reverse causality and is necessary for dynamic causal inference methods like difference-in-differences.

For these reasons, the growth of the availability of big datasets with over-time variation, such as datasets derived from social media, digitized historical documents, or long-running surveys, offer important new areas of application of the ideal point model. However, it can be difficult to repurpose ideal point models to new time-varying domains as over-time variation poses significant hurdles when trying to identify a noisy latent trait. Many advances in ideal point modeling remain limited to static models, such as missing-data adjustment [@rosasNoNewsNews2015] and non-binary outcomes [@goplerudMultinomialFrameworkIdeal2019], while available dynamic models offer a limited range of time-series processes that may not apply well to diverse phenomena [@quinn2002]. These limitations are particularly unfortunate for measurement models because the aim is to remove measurement error from available data; if the measurement model imposes additional unrealistic assumptions about the nature of the latent concept, it may result in worsening measurement error via mis-specification.

I argue that scholars should first define their latent construct theoretically, and then choose a parametric specification that matches these theoretical priors as close as possible. For this reason, the aim of this paper is to derive novel parameterizations of ideal point models that extend the framework to novel forms of latent variables and related measures. The types of data that social scientists increasingly use require diverse statistical distributions, endogenous adjustments for non-ignorable missing data (i.e., selection bias), a wide variety of possible time series processes, and robustness to both computational complexity and high-dimensional inference–requirements that are not trivial even with recent advances in statistical computation.[^1]

[^1]: Before beginning the model definition, I first note that I will only consider the case of a 1-dimensional dynamic latent variable. The reasons for this focus is largely practical; in the dynamic setting, multi-dimensionality is a less important feature to consider as a time-varying latent variable is already two-dimensional by definition (i.e., time is the second dimension). Furthermore, in social science settings, the dimensionality of latent variables is usually quite low and adding in additional dimensions can lead to overly-complex models that are difficult to interpret, especially in the case of time-varying latent variables. In any case, all of the analyses that I present below can be straightforwardly modified to accommodate additional noncompensatory dimensions.

## The Modular Approach of `idealstan`

The `idealstan` model achieves a higher level of flexibility and robustness by fully exploiting computational and statistical advances in Bayesian modeling. `idealstan` is built on top of Stan, which is a powerful sampler for Bayesian inference using Hamiltonian Markov Chain Monte Carlo (HMC) [@betancourt2013]. Relative to existing Bayesian frameworks, such as Metropolis Hastings or Gibbs samplers, HMC is designed for high-dimensional inference (i.e., lots of correlated parameters), and with an auto-differentiation library, can incorporate most statistical distributions for parameters without requiring conjugacy (i.e., mathematical tractability of the posterior distribution). Importantly, HMC and other Bayesian estimation algorithms derive uncertainty directly from the model by estimating the joint posterior distribution, which provides strong finite-sample uncertainty for all parameters.

For these reasons, Stan-based measurement models have already made a significant impact in the discipline [@barbera2015; @caugheyDynamicEstimationLatent2015; @bølstad2023]. However, it is important to note that Stan is not a model but rather an estimation engine. As such, each new specification written in Stan must be rigorously tested as with any other new model. In addition to offering new ideal point specifications, `idealstan` embeds these new models within a meta-measurement framework so that these new specifications can be combined with what is already available in the ideal point literature so as to permit researchers to select the exact model that fits their measurement problem.

<!--# By contrast, applications of Stan's facilities to a broader set of statistical problems, such as with the groundbreaking `brms` package [@burknerBrmsPackageBayesian2017], can radically increase possibilities for Bayesian model estimation and the development of robust statistical models. Furthermore, Stan employs a probabilistic programming language, which allows for models to be symbolically defined in mathematical notation [@CarpenterGelmanHoffmanEtAl2017]. While the difference from conventional programming may appear trivial, this formalization of statistical programming along with the flexibility of Stan's HMC engine means that it is possible to fit a far greater variety of models within the same mathematical framework. So far, political scientists have employed Stan to create novel methods that address difficult topics in measurement . Yet even these impressive contributions have only scratched the surface of Stan's capabilities because they have resulted in bespoke models for the relevant application. -->

For these reasons, I focus on those contributions of this measurement framework that advance the research agenda of dynamic measurement design. Each of the specifications that I present below can be understood as its own model, but I define them separately before showing how they can be jointly estimated for optimal flexibility.

## Model Definition

We can begin with the common situation in which a researcher is interested in measuring a time-varying latent concept like democracy, ideology, or state capacity. We can denote this latent concept as $\alpha_{it}$ for each unit $i$ and time point $t$, which by convention we can call an ideal point, although the trait does not need to represent political ideology. While we cannot directly observe $\alpha_{it}$, we do have a set of measurements $Y_{ijt}$ for a given set of items $j \in J$, or attributes, that represent the visible manifestation of the latent trait. We consider each item $j$ to have its own position in this latent space, which we can measure via a discrimination parameter $\gamma_j$ and difficulty parameter $\beta_j$. Items with higher discrimination $\gamma_j$ should, on average, contain more information about the latent trait $\alpha_{it}$. It should be noted that only the latent trait varies over time; an item may have repeated measures at different time points but its discrimination and difficulty parameters remain constant.

We wan to learn the posterior distribution of this set of unobserved parameters $\{\alpha_{it},\gamma_j,\beta_j\} \in \theta$ that is conditional on the observed measurements $Y_{ijt}$, or $Pr(\theta|Y_{ijt})$. To do so, I define independent priors on the set of parameters $\theta$, $Pr(\theta)$, which is multiplied by an appropriate likelihood $L(\cdot)$ for the measures $Y_{ijt}$ as shown in @eq-Bayes:[^2]

[^2]: I use the proportional symbol $\propto$ to reflect the fact that the term in the denominator representing the probability of the data $Pr(Y_{ij})$ is normally omitted in computational Bayesian analysis as it is fixed with respect to the parameters.

$$Pr(\theta|Y_{ijt}) \propto Pr(\theta)L(Y_{ijt}|\theta)$$ {#eq-Bayes}

Given the standard definition of an ideal point model as a form of an item-response theory specification [@bafumiPracticalIssuesImplementing2005], we can estimate the time-varying ideal points $\alpha_{it}$, discrimination parameters $\gamma_j$, and difficulty parameters $\beta_j$ given our observations $Y_{ijt}$ for each person $i$, time point $t$, and indicator $j$ with the following likelihood $L(\cdot)$:

$$L(Y \mid \theta) = \prod_{i=1}^{I}\prod_{t=1}^{T}\prod_{j=1}^{J}
p_{ijt}^{\,Y_{ijt}}\,(1 - p_{ijt})^{\,1 - Y_{ijt}},
\qquad
p_{ijt} = logit^{-1}(\gamma_j \alpha_{it} - \beta_j)$$ {#eq-basic}

We begin with the assumption that $Y_{ijt}$ is distributed Bernoulli and requires a logit link function as is customary in the literature.

Given that @eq-basic has no data on the right-hand side, we must impose some restrictions or priors to identify the parameters [@bafumiPracticalIssuesImplementing2005]. In a Bayesian context, we assign informative prior distributions to identify the scale and rotation of the latent traits/ideal points $\alpha_{it}$. The model presented in this paper employs an informative prior distribution (the generalized beta distribution) over items to make it easier to identify time-varying latent traits; this method is discussed in SI section 6.

It should be noted that `idealstan` also implements likelihoods for the Normal distribution, the log-Normal distribution, the Poisson distribution, ordinal outcomes for both rating scale and graded response formulations, and the ordered beta distribution for bounded continuous responses like percentages and proportions [@kubinecOrderedBetaRegression2022]. Furthermore, the modeling framework allows for these distributions to vary at the item (indicator) level and can be jointly estimated with all of the time series processes and missing-data adjustments discussed previously. For reasons of brevity, I refer the reader to SI Section 1 for the full derivation of these additional likelihoods.

## Time-varying Inference

While there is ample discussion of the interpretation of ideal point models in the canonical static context, there is much more ambiguity about the way that dynamic ideal points should be modeled. What is the correct specification to employ if a latent trait, $\alpha_{it}$, varies over time? The answer provided in this paper is that there is no single such time process; different time processes are more or less useful depending on the particular nature of the theoretical construct [@box-steffensmeier2014]. In particular, the scholar needs to consider how the theoretical construct is expected to change over time [@reuning2019] in order to select an appropriate time series specification. As a consequence, the modeling approach presented in this section focuses on implementing a wide array of time series methods from the simple to the complex in order to permit maximum comparability and compatibility with theoretical requirements.

I include in this section two time series specifications that have been employed in modeling approaches to date, though rarely together in the same model. The first method, polynomials or splines, were used by the DW-NOMINATE model [@rosenthal2007], which has a linear time trend to allow for gradual over-time variation at the legislator level across Congressional sessions. @bailey2007 later employed a 4th-order polynomial to provide common space scores for the American judiciary, which he argues, "represents a trade-off between flexibility and computation" given the sparseness of judicial decisions.

The other widely-used approach is the random walk model of @quinn2002 in which ideal points in time $t$ are equal to ideal points in $t-1$ plus a Normally-distributed random jump. This much more flexible model—at additional the cost of estimating one latent trait per person per time point—is quite helpful at determining the probable trend in ideal points over time because it imposes virtually no structure on the time-series trends compared to a simpler polynomial specification a la DW-NOMINATE and @bailey2007. However, as I show in the empirical application, the additional parameters require substantial data to estimate the trends without structure, and are therefore a difficult specification to use for sparse data structures.

While I implement both of the methods above—including a generalized parameterization for polynomial splines—I also implement two entirely new specifications for latent measures, the autoregressive equation of order 1 (AR(1)) and the Gaussian process. These two time-series specifications have not been employed to date in the dynamic ideal point literature, and I incorporate them because they increase by a considerable margin the possibilities for the type of dynamic latent construct that can be estimated. An AR(1) time-series variable will always return to an over-time mean, i.e, it is "stationary." By contrast, a Gaussian process is one of the most flexible kinds of time-series processes, permitting dramatic shifts from one time point to the next—at the cost of considerably more computational complexity than the random-walk parameterization.

I do not argue in this paper that one of these time-series specifications is superior across all contexts. Rather, I aim to show how these different processes can solve different measurement challenges given costs and trade-offs in terms of the availability of prior information and, most importantly, the known nature of the latent trait.

I begin by defining the random walk as a relationship in the prior between $\alpha_{it}$ and $\alpha_{it-1}$:

$$\alpha_{it} \sim N(\delta_i+ \alpha_{it-1},\sigma_i)$$

We further give every person $i$ a separate over-time variance parameter $\sigma_i$ and intercept/offset $\delta_i$. For computational reasons, the implementation of this model in `idealstan` employs a non-centered parameterization [@betancourt2013] that reduces dependence between the variance $\sigma_i$ and the prior value of $\alpha_{it-1}$:

$$\begin{split}
\epsilon_{it} &\sim N(0,1)\\
\sigma_i &\sim E(1)\\
\alpha_{it} &= \delta_i + \alpha_{it-1} + \sigma_i\epsilon_{it}
\end{split}$$

The first new time-series model that I consider is an AR(1) or autoregressive parameterization, which can be understood as a generalization of the random-walk model. To do so, we must add a parameter to the model, $\psi_i$ that is constrained to lie in the $(-1,1)$ interval:

$$\begin{split}
\epsilon_{it} &\sim N(0,1)\\
\sigma_i &\sim E(1)\\
\alpha_{it} &= \delta_i + \psi_i\alpha_{it-1} + \sigma_i\epsilon_{it}
\end{split}$$

If $\psi_i$ lies in the $(-1,1)$ interval, then the series is stationary and will always return or decay to the long-term mean of the series, $\dfrac{\delta_i}{1-\psi_i}$. By comparison, the random walk model can be thought of as an AR(1) model where $\psi_i$ is fixed at 1. This type of time series model implies that a given person or legislator has a stable over-time average ideal point and that movement away from that over-time average is some kind of exogenous shock that will eventually decay.

Both the AR(1) model and the random walk are models of discrete time, or a set number of time points. The next two time series processes that I consider are both continuous models of time; given any set of time points, they are capable of producing smooth interpolations between observed time points. The first such time process I consider is the spline method of constructing polynomials.

Splines are a method of constructing semi-parametric time-series by combining polynomial functions of time [@ratkovic2010]. While there are many different kinds, I consider here basis splines due to their ability to fit into the latent variable framework. A basis spline separates a polynomial function of time into individual basis functions that the space of all possible polynomial functions for a given polynomial degree. Because each of the individual functions are relatively simple, this construction allows for a wide variety of polynomial functions to be derived while also enabling efficient sampling.

One central feature of splines is that they are polynomial functions defined over a series of sequential partitions of a given time series known as knots. A spline of a given order is a polynomial function of a given degree (by convention, one less than the order) that is defined for each sequential partition given by the location of the knots. Each piecewise polynomial function must begin and end at the same point of each sequential partition, ensuring a smooth combined function.

The presentation below follows that of @kharratzadeh2017. Given a spline order $d$ and a given number $s$ of sequential knots $q$, we can define a basis matrix $B_{s,d}$ as a function of continuous time points $t \in \mathbb{R}|t_{min} < t < T$:

$$B_{s,d}(t) = \omega_{s,d}\, B_{s,d-1}(t) + (1 - \omega_{s+1,d})\, B_{s+1,d-1}(t)$$ {#eq-spline1}

where the function $\omega(\cdot)$ can be defined as:

$$\omega_{s,d} = \begin{cases} 
\frac{t-q_s}{q_{s+d-1}-q_s}, & \text{if } q_s \neq q_{s+d-1} \\
0, & \text{otherwise}
\end{cases}$$ {#eq-spline2}

The resulting basis matrix $B_{s,d}(t)$ is a set of vectors of number of columns $T$ and number rows equal to the degree of the polynomial plus the number of knots. The polynomial function of $t$ can then be generated by multiplying the basis matrix by a vector of spline coefficients $A_i$ :

$$S_{q,d}(t) = B_{s,d}(t)A_i$$

where the length of spline coefficient vector $A_i$ is equal to the number of knots plus the degree of the spline. While in practice $t$ is in a set of discrete time points, defining the function over a continuous interval permits interpolation of the time series, a convenient feature of the spline technique.

We can then sample the spline coefficients by defining a loose prior distribution for $A_i$ as a function of a scaling parameter $\tau$ and a location parameter $A'_i$:

\begin{align}
A'_i &\sim N(0,1)\\
\tau &\sim E(1)\\
A_i &= A_i'\tau
\end{align} \label{eq-splineprior}

The adjustment with $A_i'$ and $\tau$ is to permit non-centered sampling which is more efficient.

The main advantage of using basis splines as I demonstrate in the empirical application is that it is possible to fine-tune the complexity of the time series function, which is valuable as it allows for theoretical priors to govern the amount of possible over-time variation in the latent trait. Increasing either the degree of the polynomial or the number of knots increases the number of coefficients per person by one. In cases of high sparsity, a low number of knots (or even no knots at all) and a low degree polynomial can be used to find a stable latent variable estimation even with very limited data per time point. This point was first made by @bailey2007 and, I argue, is a major reason why splines can be a useful way to incorporate theoretical priors into ideal point estimation.

Finally, a Gaussian process (GP) for the ideal points $\alpha_{it}$ represents the most flexible framework possible for semi-parametric inference of either spatial or temporal autocorrelation [@rasmussen2006]. Given the limited application of this model in political science research (though see @duck-mayr2020 for an application to item-response theory utility functions), I briefly review the notation for a GP before combining it with the ideal point model.

A GP is usually defined as a "distribution over functions" [@rasmussen2006, 13]. We assume that underlying an observed time series $x_t$ is an unobserved function $f(x_t)$. To estimate the most likely values of $f(x_t)$, $f(x_t)$ is defined as a multivariate Normally-distributed latent variable where the mean and covariance of this distribution are themselves functions of the mean and covariance of $f(x_t)$:

$$f(x) \sim N\big(\mu(x),\, K(x,x')\big), \qquad [K(x,x')]_{t,t'} = k(t,t')$$ {#eq-gp}

Usually the $\mu(f(x_t))$ function is assumed to be 0 as it does not vary with $t$. Instead, much of the flexibility of the distribution involves specifying a function for $K(x,x')$, the covariance. The function most often used, and incorporated here, is the squared-exponential kernel $k(t,t')$:

$$k(x_t,x_{t'}) = \large \sigma_f  e^{\displaystyle -\frac{(x_t - x_{t'})^2}{ 2l^2}} + \sigma^2_{i,n}$$ {#eq-gpcov}

where $\delta_{t,t'} = 1$ if $t = t'$ and $0$ otherwise.

What this function does is convert all the distances in terms of time between $x_t$ and $x_{t'}$ into a positive semi-definite covariance matrix $\Sigma$. We can then sample from this multivariate Normal with covariance $\Sigma$ to estimate a distribution over the possible time-series functions $f(x_t)$. The GP framework is so flexible that it can fit an *infinite* number of basis functions of $x_t$ [@rasmussen2006, 14]. Other semi-parametric functions, such as splines and polynomials, are special cases of the GP for certain values of the hyper parameters used in the covariance function [@rasmussen2006, 137-140]. What is even more appealing is that the GP can handle time-varying covariates $X_t\phi'$ simply by including them in place of $\mu(x_t)$. The multivariate Normal will then sample from the specified covariate matrix while averaging over the possible values of the covariates.

What gives the GP the ability to fit so many different functions are the three hyper-parameters in @eq-gpcov. $\sigma^2_{f}$ can be referred to as the marginal standard deviation and represents the total amount of variance in $x_t$ explained by the covariance function $k(x_t,x_{t'})$. A higher value for this hyperparameter will result in more bounce in the time series. $\sigma^2_{i,n}$, on the other hand, represents residual variance in the time series $x_t$ that the covariance function does not fully explain (which could be interpreted as additional measurement error). Finally, the length-scale parameter $l^2_i$ represents a smoothing factor controlling how much different in time are correlated together. A very low length-scale will allow the time-series to cross the origin at a much higher rate. Conversely, higher length-scales result in smoother functions. The three hyper-parameters both overlap and interact, which makes isolating the effect of any one hyper-parameter difficult.

To employ the GP as a time process for ideal points, we use the time indices $x_{t}$ (which do not vary across persons) as the input to the kernel function $k(\cdot)$ with the latent ideal points $\alpha_{it}$ as the output:

$$\alpha_{i,\cdot} \sim N(0,\, K_i), \qquad [K_i]_{t,t'} = k(t,t')$$ {#eq-gprior}

For time periods $t$ for which no votes are observed, either the missing-data inflation mechanism can be used as described elsewhere or the the ideal point $\alpha_{i,t}$ will simply default to the prior distribution.

This straightforward parameterization shows some of the power of Bayesian modeling. We can include virtually any time process by simply changing the definition of the prior distribution for the ideal points $\alpha_{it}$. Of course, different time series processes for $\alpha_{it}$ have varying levels of required information to find unique solutions.

In summary, these different time series models can be understood in terms of order of complexity. Except for splines, all of the time series models require at least one parameter per time point $t$. The random walk is the simplest of these models as it only requires one additional parameter, $\sigma_i$, per person $i$. The AR(1) model is more complex as it introduces an additional parameter to control the decay rate in the series, $\psi_i$. The GP is the most complex as it has three separate parameters per person, $\sigma^2_{f}$, $l_i^2$ and $\sigma^2_{i,n}$, in addition to one estimate per time point $t$. For these reasons, the GP is the most demanding model to fit for a latent variable model as there needs to be a substantial amount of data per time point to estimate a flexible time trend while also separating out measurement noise.

For these reasons, I argue that lower-dimensional time series methods like splines can be useful even given that functions like the GP (or even a random walk) are by nature more flexible. As the ideal point model is a measurement model, it can be difficult to separate measurement noise from the additional noise arising from flexible time series processes. In order for measurement to be successful, the time series specification should match the theoretical understanding of the latent construct of interest. I will illustrate these trade-offs later in this paper in the empirical section.

## Time-Varying Non-Ignorable Missing Data

Missing data is a tricky problem to address in latent variable models because latent variables are themselves defined as missing data. For this reason, a straightforward approach of imputing data conditional on the model assumes that missingness is orthogonal to observed ideal points [@rubin2002]. In many situations, this assumption may not be realistic when there is reason to believe that social actors, such as legislators, do not respond to stimuli out of a desire to avoid revealing their latent trait. In this section I define a model for non-ignorable missingness in ideal points that can be integrated with time-varying processes (and of course applies to static models as well).

It is important to first note that Bayesian variants of the ideal point model, such as @jackman2004, are already able to incorporate ignorable missingness in dynamic models because the model can default to the prior in the case of missingness. That is, suppose for some time series of length $T$ that a legislator fails to vote at $t=4$ and $t=5$. In this case, a properly specified Bayesian model can still estimate legislator trajectories for $t=4$ and $t=5$ conditional on the prior (i.e., time-series process) and the legislator observed votes for all other time periods. Crucially, however, this mechanism can be classified as "conditionally missing at random" (CMAR) per @rubin2002 terminology because all information about the nature of missingness must be in the observed votes or in the prior. Given that most Bayesian model use weakly or non-informative priors, this assumption essentially amounts to exogenous missingness conditional on pre- and post-trends. However, if missingness is endogenous to the ideal point score at the missing time point, such as a legislator refusing to vote on a certain bill because they are liberal or conservative, than the missingness process is "missing not at random" (MNAR) per @rubin2002 and bias necessarily results, as I will show later in the simulation results.

At present, there are relatively few discussions of non-ignorable missing data in the ideal point literature, in part because non-ignorable missingness in latent variable models is a tricky problem. The most robust discussion of this problem to date is in @rosasNoNewsNews2015 who derive an extension in which legislators decide not to show up to vote if their ideal point differs from their party's ideal point. This model turns the missing data problem on its head by deriving interesting empirical findings from missingness rather than attempting to impute it away. I adopt the insight of this model by focusing on endogenizing missingness as a way of learning about ideal point distributions.

In the general IRT literature, there has been a more extensive discussion of non-ignorable missingness in terms of students failing to complete all test questions due to time constraints [@mislevy2016; @ulitzsch2020]. @holman2005 derived a general model of missingness in IRT by allowing for separate ability scores for observed and missing data where the scores are related to each other via a multivariate Normal distribution. These approaches are not as easy to apply to ideal point models because traditional IRT models make the assumption that positive item scores are always associated with higher ability scores, while the ideal point model allows for some items to negatively predict the latent trait (such as liberal and conservative bills in a legislature). Consequently, an appropriate missingness adjustment for an ideal point model must also allow for non-ignorable missingness to affect both poles of the latent trait (such as liberal versus conservative abstention patterns).

The approach that I present here bears similarity to these other approaches in that it jointly estimates a set of latent missingness item parameters with the observed latent item parameters. The relationship between the missing and observed latent factors is endogenized by adopting a selection approach, specifically, a hurdle model, as I show below. Given the issues of identifying time-varying latent variables that I discuss later, the proposed missingness model allows for substantial inflation or deflation of the latent trait due to selection bias while still remaining tractable for dynamic ideal point estimation and identification. While it is impossible for a single model to embody *all* potential forms of MNAR ideal point data, this general selection model addresses one of the main forms of strategic selection into measurement as a function of unobserved ideal points: do people who align with one end of the ideological spectrum tend to select into measurement (such as casting votes) at a higher rate than people of opposite ideological persuasion?

To define the missingness hurdle extension, I extend my notation of the outcome $Y_{ijt}$ by adding a subscript $r \in \{0,1\}$ for whether person $i$ chose to vote or answer item $j$ ($r=1$) or chose not to answer ($r=0$). We can then add a separate selection model that first estimates $Pr(r=0)$ and deflates the likelihood $L(Y_{ijtr}|\theta)$ accordingly:

$$L(Y_{ijtr} \mid \alpha_{it},\gamma_j,\beta_j,\nu_j,\omega_j) =
\prod_{i=1}^{I} \prod_{t=1}^{T} \prod_{j=1}^{J}
\begin{cases}
\zeta(\alpha_{it}\nu_j - \omega_j) & \text{if } r=0 \\[2mm]
\big(1-\zeta(\alpha_{it}\nu_j - \omega_j)\big)\, p_{ijt}^{\,Y_{ijt}}(1-p_{ijt})^{1-Y_{ijt}} & \text{if } r=1
\end{cases}$$ {#eq-inflate2}

where $\zeta(\cdot)$ is the inverse-logit function and $p_{ijt} = logit^{-1}(\gamma_j\alpha_{it}-\beta_j)$ as in @eq-basic.

I let $\zeta(\cdot)$ stand for the inverse logit function in the equation above. As can be seen, the selection model for missing data $\zeta(\alpha_i'\nu_j - \omega_j)$ is very similar to @eq-basic except that I have substituted a new set of discrimination $\nu_j$ and difficulty $\omega_j$ parameters. The ideal points $\alpha_{it}$ enter into both the selection model and the main ideal point model $L(Y_{ijtr}|\alpha_{it},\gamma_j,\beta_j)$. As such, the selection model is able to inflate or deflate the ideal points by taking into account a first stage process. The inclusion of $\nu_j$ and $\omega_j$ in the first-stage selection model creates a new "missingness" space where person $i$ first chooses whether she will cast a vote, send a tweet or answer a survey question, as the case may be. Only if the item is close to person $i$ in this missingness space will person $i$ then also decide to participate and provide information $Y_{ijtr=1}$ that is informative of their ideal point. This model can be interpreted as a censoring model where persons may choose to self-censor their ideal points.

The general form of the hurdle model--which is essentially a first-stage ideal point model--allows it to pick up a range of non-ignorable missingness patterns if missingness correlates with ideal points. The model does not make a priori assumptions about exactly why ideal points may determine missingness, and as such the model cannot provide such an interpretation a posteriori without taking time to interpret what the parameters of the selection model reveal about actors' intentions. The missingness discrimination parameter $\nu_j$ is very helpful for this purpose because it will indicate which set of items show correlation between missingness and one of the poles of the latent variable. For example, if the latent variable is constrained positive for liberal Senators, then high positive discrimination would indicate that more liberal Senators tend to be absent on a particular bill. A naive model that assumed that absences were ignorable would miss this pattern and treat these Senators as more moderate based solely on their observed voting record.

This model is also general enough to allow for the possibility that data is actually missing at random at the item level. If the discrimination parameter $\nu_j$ is zero then ideal points do not enter into the equation for missingness and $Pr(r=0)$ is equal to the item-specific intercept $\omega_j$. In the legislative context, this would be the same as estimating the probability of absence by the proportion of legislators who do not show up for particular bill. Including this parameter will also separate that item-specific random missingness from missingness patterns suspiciously associated with ideal points. For this reason, the selection model will perform at least as well as a standard imputation method where the missing ideal points are MCAR conditional on each item.

While this missingness model is readily applicable to a legislative context where legislators may not want to show up on votes depending on what the vote would reveal about their ideal points, it is also useful for other non-ignorable missingness patterns. Twitter data is an excellent example. It is well-established that estimating a latent trait like political polarization will be vastly over-stated if the tendency of Twitter (now known as X) users to select who they choose to follow is not taken into account [@Barbera2015]. The two-stage selection model can be used to estimate the people's propensity to retweet ideological content net of their self-selection into whom they follow on social media [@kubinecWhenGroupsFall2021]. For survey data, the two-stage model will take into account that respondents with high or low values on the latent trait could be more or less likely to provide an answer on a given question, and then backwards infer their true ideal point given those varying missingness rates by question.

## Big Data Inference

The Achilles heel of Bayesian inference is generalizing to large data sets. Markov chains, which form the core of the estimation engine, cannot be easily parallelized as each iteration depends on the value of the previous iteration. However, recently the Stan team has implemented a way to parallelize the *posterior gradient* computation within a Markov chain, which are necessary and very expensive to calculate the iteration trajectories of Hamiltonian MCMC. Any model likelihood can be parallelized in terms of its gradient calculations so long as the parameters are conditionally independent within discrete subsets or "shards" of the data.

The `idealstan` model solves the independence problem by calculating likelihoods within discrete subsets of parameters. For a static ideal point model, this constraint means that the likelihood is parallelized across subsets of persons or items. For a dynamic model, parallelization is performed across persons as the ideal point for a given person at time $t$ depends on the value of the ideal point in $t-1$, which would break conditional independence of parameters if the data were subset by item.

This type of parallelization of the gradient calculations is implemented for all of the models previously described. This allows for the computation of Bayesian ideal points for datasets that were previously intractable due to the length of time required. However, it should be noted that this type of parallelization does have significant overhead as not all of the computations of the Markov Chain are parallelizable, and as a result significant information must be shared across parallel threads.

Even with these restrictions, though, this type of computation greatly expands the possibilities for Bayesian inference on big datasets, reducing the time required for model convergence by several orders of magnitude depending on the number of cores available. In the supplementary information section 3 I show a simulation of `idealstan` models of varying sizes and the associated speedups from using multiple cores. Moving from a single core to four cores, which is available on many laptops, can cut estimation time for a medium-sized dataset from approximately 20 minutes to approximately 7 minutes. With larger numbers of cores (up to the limit of what is available in single machines in conventional computing clusters), run times can be reduced to as low as a single minute for datasets of up to 100,000 observations.

In addition, variational methods that approximate posterior distributions can be used as well in place of MCMC for the largest datasets [@Grimmer2011]. @imaiFastEstimationIdeal2016 were the first to apply variational methods to ideal point estimation, demonstrating that the tool could achieve remarkable gains from speed. However, @imaiFastEstimationIdeal2016 had to derive a separate variational algorithm for each model, restricting the application of variational approximations to binary data in the case of dynamic ideal point modeling.

`idealstan` implements two variational methods that integrate with Stan and allow for all of the time-varying methods mentioned previously to be estimated. The first is the Pathfinder method described previously, which can also be used for final estimation in addition to finding starting values for MCMC. The second is the Laplace approximation of the joint posterior distribution [@rue2009], which estimates the mode of the posterior distribution with optimization and then uses the Normal distribution to calculate uncertainty. The `idealstan` model can estimate all of the previously described models with these approximation methods.

Both of these methods are implemented in `idealstan` as both approximations may have utility depending on the particular measurement problem and the extent of bias caused by using an approximation to the true joint distribution. However, given that much is unknown about the performance of these algorithms with latent variable models and dynamic ideal point models specifically, I test them against MCMC in the simulation I describe in the next section.

## Monte Carlo Simulation

Using the specifications described previously, I generated simulated data with a Bernoulli (binary) outcome for dynamic ideal points $\alpha_{it}$ with each of the time series processes (splines, GPs, random walk, and AR(1)). I fix the number of persons in the model at 30 and vary the number of items from 100 to 400. I also vary whether non-ignorable missing data is present or absent, and the number of time points from 10 to 20. The data are generated using the same priors described previously except that I use an informative $\text{Normal}(0,1)$ prior on the initial time period ideal points to match existing specifications. I do not consider other outcome distributions as there are no comparable models for dynamic ideal points with non-binary distributions at present.

I ran the simulation using `R` as `R` contains the best coverage of ideal point models. Alternative specifications that I tested included `mcmcPACK` [@quinn2002], `emIRT` [@imaiFastEstimationIdeal2016], and DW-NOMINATE [@carrollMeasuringBiasUncertainty2009a].[^3] Output from these models was standardized using the posterior draws in the case of `mcmcPACK` and using parametric bootstrap draws in the case of `emIRT` and DW-NOMINATE. I also include the Pathfinder and Laplace approximation methods in addition to Hamiltonian Markov Chain Monte Carlo in `idealstan` as separate estimates. `idealstan`'s HMC method is allowed to use 4 cores to speed up estimation, which is also provided to `emIRT`. I used default settings for estimation, and for `idealstan` and `mcmcPACK`, I ran their respective samplers for 1,000 and 60,000 iterations (HMC samplers require fewer iterations than Gibbs samplers). I also set priors where available to match `idealstan` priors, i.e. such as tight $N(0,1)$ priors on ideal point intercepts.

[^3]: For DW-NOMINATE I employed the R package `dwnominate` , which is a front-end to the original FORTRAN code. See <https://github.com/wmay/dwnominate>.

I also generate an additional observed outcome $Y$ using simulated ideal points $\alpha_{it}$ via the following specification:

\begin{align}\label{eq-sim1}
X_{1,it} &\sim \text{Normal}(1 + 0.5\,\alpha_{it},\, 1)\\
Z_{it} &\sim \text{Normal}(-2 + \beta_S\,\alpha_{it} X_{1,it} + 3 X_{1,it},\, 1)
\end{align}

where $\beta_S$ is the coefficient of interest for recovery and fixed at +0.025. This particular specification is employed to correspond to "real-world" application of ideal point models in which the estimated scores are used as covariates in regression modeling. The endogenous and interactive nature of covariate $X_1$ is used to make inference more challenging.

The aim of the simulation is to compare the different models by how well they can estimate the true (standardized) ideal points, measured with RMSE and Kendall's tau rank coefficient for rank order, along with inference on downstream quantities ($\beta_S$). I also examine coverage in the SI (see section 5), but I do not further discuss those results as comparing coverage across these models is very tricky given that the approximation methods such as parametric bootstrapping and `idealstan`'s Pathfinder and Laplace generally do not capture full uncertainty. Instead, focusing on RMSE and Kendall's tau provides a comparable metric across methods given that it is quite difficult to capture "true" values for latent quantities.

For the latent regression simulation, I measure downstream inference performance using the S error and M error statistics in which an S error occurs when an estimate of $\beta_S$ is statistically significant at the 0.05 level but of the wrong sign and an M error is the relative inflation of statistically significant estimates to the true values (i.e., the absolute value of the ratio) [@gelman2014types]. I employ these measures, as opposed to predictive measures of model fit, as they correspond more closely to the goals scholars aim to achieve with ideal point models: measuring the latent trait accurately for a set of persons or institutions and using estimated scores as covariates in downstream analyses.[^4] Given that the true effect $\beta_S$ is quite small and there is significant measurement error, the S and M error rates should be substantial for all models.

[^4]: Some scholars have been employing the method of composition to propagate measurement errors in downstream models [@jerzak2024], but lately this method has come under criticism for lacking a clear statistical basis. For these reasons I do not perform this additional adjustment as error propagation is not particularly important for the task of comparing across models.

Because estimation of these models can be quite lengthy—especially for Markov Chain Monte Carlo methods like `mcmcPACK`--I have to limit the number of draws of the simulation. Employing a high-performance cluster [@calderOokamiA64FXComputing2024], I was able to produce 5,760 simulation draws across 288 separate conditions. Because each simulation had a large number of ideal point parameters, some statistics like RMSE, which are defined relative to the full set of generated ideal points, are estimated more precisely than others that only varied across simulation draws like M and S errors of downstream regressions.

All models are given the same amount of identification information (persons or item parameters), and in addition, if a model converges to the wrong mode (i.e., the sign of the ideal points is flipped), the ideal points are sign-reversed before calculating relevant statistics to avoid obfuscating model fit with arbitrariness in identification conditions. While sign-identification can vary due to mis-specification, that issue is beyond the scope of these simulation results as it adds an additional layer of complexity to model comparison.

# Results

In this section I first discuss model comparison with the simulation results and then turn to an application of `idealstan` to the estimation of Congressional dynamic ideal points.

## Simulation

The simulation results for latent trait recovery are shown in @fig-rmse and @fig-kendall and the latent regression results in @fig-serrs and @fig-merrs. @fig-time shows `idealstan` performance in terms of estimation speed. SI Section 5 contains results for coverage and sign-identification across models.

The latent trait recovery results for RMSE (@fig-rmse) and rank order (@fig-kendall) shows that `idealstan` with HMC has a lower RMSE and a higher Kendall tau across time processes, as would be expected given that `idealstan` is the "true" model. While all models have significant residual error--as is expected for any latent measurement model--`idealstan`'s performance is measurably better than alternatives. Furthermore, the missing data adjustment has almost no penalty for `idealstan`, showing that `idealstan` with the HMC estimation method is able to fully endogenize this missingness and return accurate measures of ideal points. As can be seen, there are considerable penalties for other models' estimation accuracy when either the time process is mis-specified, missing data is ignored, or both in combination.

```{r fig-rmse}
#| fig-cap: "RMSE of Estimated to True Ideal Points by Model"
#| fig-height: 4
#| label: fig-rmse

rmse_plot <- readRDS("data/plot_rmse_all_corrected.rds")

rmse_plot

```

```{r}
#| fig-cap: "Kendall Tau Rank Coefficient of Estimated Ideal Points by Model"
#| label: fig-kendall
#| fig-height: 4

kendall_plot <- readRDS("data/kendall_tau.rds")

print(kendall_plot)
```

```{r serrs}
#| fig-cap: "S Error Rates for Downstream Regression of Estimated Ideal Points by Model"
#| cache: false
#| label: fig-serrs
#| fig-height: 4

s_err_plot <- readRDS("data/s_errors.rds")

s_err_plot

```

```{r merrs}
#| fig-cap: "M Error Rates for Downstream Regression of Estimated Ideal Points by Model"
#| cache: false
#| label: fig-merrs
#| fig-height: 4

m_err_plot <- readRDS("data/plot_m_errors.rds")

m_err_plot

```

It is very interesting to compare the performance of the approximation methods and alternative specifications. `emIRT` and `mcmcPACK` show significant declines in performance when missing data is included in both RMSE (@fig-rmse) and Kendall tau coefficients (@fig-kendall), though the difference is more noticeable in the Kendall tau coefficients. The approximation methods Laplace and Pathfinder were more robust than the non-`idealstan` methods without missing data, although the approximations performed better in simpler time series processes like splines and random walks. On the other hand, a noticeable gap exists for the approximations relative to full HMC in the Gaussian process results.

`mcmcPACK`, which is the other Bayesian method under consideration, performed best in the random walk specification, which would be expected as that is the same time series process it employs. `mcmcPACK` performed as well as `emIRT` on the Gaussian process and AR(1) specifications, but performed much more poorly on the spline method. In fact, `emIRT` performed quite well on the spline process, with its RMSE score better than the Pathfinder and Laplace approximation methods even though it employs a random walk time series specification. This exceptional performance might be a side effect of the variational method in `emIRT` that simplifies the random walk process into a lower-dimensional space [@imaiFastEstimationIdeal2016].

The results for the downstream regression of estimated ideal point scores show similar trends with some noticeable differences. @fig-serrs reveals that the difference between models is more modest for the S error metric (obtaining a latent regression coefficient of the wrong sign), possibly because S error rates are on the whole relatively low even with a small true coefficient size. However, there is a noticeable difference between estimations with missing and non-missing data for the random walk, AR(1), and Gaussian process specifications, with methods like `mcmcPACK` resulting in S error rates that are about 50% greater on average when non-ignorable missingness is present.

The M error plot (i.e., over-estimation of latent regression effect) in @fig-merrs shows that there is indeed a substantial penalty for alternative methods in terms of time series and missing data mis-specification. Interestingly, while the approximation methods Laplace and Pathfinder perform better on average than `emIRT` and `mcmcPACK`, they show high M error rates with missing data adjustment, while the `idealstan` method with HMC does not. As can be seen, the approximation methods only seem to perform well on simpler time series specifications without non-ignorable missingness, though they can still out-perform a robust method like MCMC with the wrong model.

Finally, @fig-time shows that `emIRT` remains the fastest estimation method as a function of the number of items included in the simulation, followed by `idealstan` with the Pathfinder and Laplace approximations, which is to be expected given that `emIRT` contains custom derived algorithms for variational approximation. However, when it comes to full Bayesian estimation, `idealstan` shows the benefits of its parallelization method. `idealstan` with HMC and `mcmcPACK` are close to each other in estimation time when the number of items is small (100), but diverge significantly as the number of items increases. Estimation time for `idealstan` with parallelization of 4 cores is approximately linear in the number of items $N$ ($O(N)$) while `mcmcPACK` and its Gibbs sampler is approximately quadratic in the number of items ($O(N^2)$), which makes larger datasets intractable for `mcmcPACK`.

The reason for the superior performance of `idealstan` with a larger number of items is that the parallel overhead remains relatively constant, which makes the `idealstan` within-chain parallelization scheme more efficient as datasets increase in size. In addition, speed can be improved for `idealstan` by increasing the number of cores up to the number of persons in the data. The level of speed enhancement for an increasing number of cores up to 48 and varying dataset sizes with `idelastan` is shown in SI Figure 2: given enough available cores, `idealstan` can obtain a full Bayesian estimation of datasets with over 50,000 observations (such as 100 legislators and 500 votes) in roughly an hour.

```{r time}
#| fig-cap: "Estimation Time by Model and Number of Items"
#| cache: false
#| label: fig-time

time_plot <- readRDS("data/plot_time_elapsed_n_items.rds")

time_plot

```

## Empirical Application

In this section, I combine the missing-data adjustment along with the new time-series methods in `idealstan` to show how to connect modeling approaches to theories. Unlike existing methods, I do not simply adopt a single time series process, but rather explore the posterior trajectory of different time series processes to determine which process most closely fits the latent concept of interest, which is in this case the canonical application of legislator ideology in the United States.

To do so, I fit time-varying ideal point models to estimate monthly legislator ideal points for Congresspeople in the U.S. House from 1990 to 2018–a 24-fold increase in resolution from the Congressional session-level dynamic ideal points that have been used for some time [@poole1997]. The increase in measurement resolution allows us to understand what might be happening *within* a Congressional session, variation that has been difficult for scholars to access. For example, @rothenberg2000a had to use a regression model to smooth the single DW-NOMINATE estimate across a Congressional session in order to understand how legislative strategic abstention changed over time. With `idealstan`, we can gain trajectories of legislator ideology that vary across and within sessions, permitting much more nuanced identification of factors that may change ideology and political affiliation over time.

However, accomplishing this task also requires selecting a theoretically-appropriate time series specification as I have discussed previously. To demonstrate how `idealstan` facilitates this task, I first present results for the 115th Congress (2016 to 2018) as it permits easier comparison of the different time series methods.

I estimate time-varying Congressperson ideal points using all four time series specifications discussed so far: random walk, AR(1), Gaussian process, and splines. For the spline specifications, I consider splines of degrees 2, 3 and 4 while using only one knot (i.e., one polynomial function for the whole time period). As I show later, when modeling multiple Congressional sessions, using knots for presidential administrations allows for period effects in the time series. With only one Congressional session, however, a single polynomial function is sufficient as it would not be expected for a member of Congress to switch their ideological position in a noticeable way more than once or perhaps twice per term. This theoretical knowledge of the latent trait is critical in order to obtain relevant measures.

To preview the findings, I show that the simpler time series models, i.e. the splines, are the most theoretically-valid for higher resolution ideal points for this type of legislative behavior. The other more complicated time-series models are only identified because of the use of Pathfinder to identify starting values in a single mode. These more flexible models are clearly not appropriate for a theoretical construct that can only vary slowly from one month to the next. The issue is not technical but rather theoretical: time-series specifications with large numbers of parameters per legislator allow more over-time change in ideal points than the theoretical construct (ideology) permits.

```{r loadmodels,include=FALSE}

if(run_all) {
  
unemp1_fit <- readRDS('data/1151_12_1_fit.rds')
unemp2_fit <- readRDS('data/unemp1151_12_2_fit.rds')
unemp3_fit <- readRDS('data/unemp1151_12_3_fit.rds')

unemp1_fitm <- readRDS('data/1152_12_1_fit.rds')
unemp2_fitm <- readRDS('data/unemp1152_12_2_fit.rds')
unemp3_fitm <- readRDS('data/unemp1152_12_3_fit.rds')

unemp_gp_fit <- readRDS('data/unemp115_1_12__gp_fit.rds')
unemp_ar_fit <- readRDS("data/unemp115_1_12_1_ar_fit.rds")
unemp_rw_fit <- readRDS("data/unemp115_1_12_1_rw_fit.rds")

unemp_gp_fitm <- readRDS('data/unemp115_2_12__gp_fit.rds')
unemp_ar_fitm <- readRDS("data/unemp115_2_12_1_ar_fit.rds")
unemp_rw_fitm <- readRDS("data/unemp115_2_12_1_rw_fit.rds")

rhats1 <- id_plot_rhats(unemp1_fit) + ggtitle("Spline 2nd Degree") + labs(caption="",x="") + xlim(c(0.95, 2))

saveRDS(rhats1, "data/rhats1.rds")

rhats2 <- id_plot_rhats(unemp2_fit) + ggtitle("Spline 3rd Degree") + labs(caption="",y="",x="") + xlim(c(0.95, 2))

saveRDS(rhats2, "data/rhats2.rds")

rhats3 <- id_plot_rhats(unemp3_fit) + ggtitle("Spline 4th Degree") + labs(caption="",y="",x="") + xlim(c(0.95, 2))

saveRDS(rhats3, "data/rhats3.rds")

rhatgp <- id_plot_rhats(unemp_gp_fit) + ggtitle("Gaussian Process") + labs(caption="") + xlim(c(0.95, 2))

saveRDS(rhatgp, "data/rhatgp.rds")

rhatar <- id_plot_rhats(unemp_ar_fit) + ggtitle("AR(1)") + labs(caption="",y="") + xlim(c(0.95, 2))

saveRDS(rhatar, "data/rhatar.rds")

rhatrw <- id_plot_rhats(unemp_rw_fit) + ggtitle("Random Walk") + labs(caption="",y="") + xlim(c(0.95, 2))

saveRDS(rhatrw, "data/rhatrw.rds")

un1 <- id_plot_legis_dyn(unemp1_fit,use_ci=F,plot_text = F,person_line_alpha = 0.1) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Spline of 2nd Degree") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))
  
  saveRDS(un1, "data/un1.rds")

un2 <- id_plot_legis_dyn(unemp2_fit,use_ci=F,plot_text = F,person_line_alpha = 0.1) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Spline of 3rd Degree") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))

  saveRDS(un2, "data/un2.rds")

un3 <- id_plot_legis_dyn(unemp3_fit,use_ci=F,plot_text = F,person_line_alpha = 0.1) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Spline of 4th Degree") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))

  saveRDS(un3, "data/un3.rds")

gp <- id_plot_legis_dyn(unemp_gp_fit,use_ci=F,plot_text = F,person_line_alpha = 0.1) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Gaussian Process") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))

  saveRDS(gp, "data/gp.rds")

ar <- id_plot_legis_dyn(unemp_ar_fit,use_ci=F,plot_text = F,person_line_alpha = 0.1) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("AR(1) Process") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))

  saveRDS(ar, "data/ar.rds")

rw <- id_plot_legis_dyn(unemp_rw_fit,use_ci=F,plot_text = F,person_line_alpha = 0.1) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Random Walk") +
  scale_x_date(guide = guide_axis(n.dodge = 2)) + labs(y="")

  saveRDS(rw, "data/rw.rds")
  
un1m <- id_plot_legis_dyn(unemp1_fitm,use_ci=F,plot_text = F,person_line_alpha = 0.1) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Spline of 2nd Degree") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))
  
  saveRDS(un1m, "data/un1m.rds")

un2m <- id_plot_legis_dyn(unemp2_fitm,use_ci=F,plot_text = F,person_line_alpha = 0.1) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Spline of 3rd Degree") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))

  saveRDS(un2m, "data/un2m.rds")

un3m <- id_plot_legis_dyn(unemp3_fitm,use_ci=F,plot_text = F,person_line_alpha = 0.1) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Spline of 4th Degree") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))

  saveRDS(un3m, "data/un3m.rds")

gpm <- id_plot_legis_dyn(unemp_gp_fitm,use_ci=F,plot_text = F,person_line_alpha = 0.1) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Gaussian Process") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))

  saveRDS(gpm, "data/gpm.rds")

arm <- id_plot_legis_dyn(unemp_ar_fitm,use_ci=F,plot_text = F,person_line_alpha = 0.1) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("AR(1) Process") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))

  saveRDS(arm, "data/arm.rds")

rwm <- id_plot_legis_dyn(unemp_rw_fitm,use_ci=F,plot_text = F,person_line_alpha = 0.1) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Random Walk") +
  scale_x_date(guide = guide_axis(n.dodge = 2)) + labs(y="")

  saveRDS(rwm, "data/rwm.rds")
  
un1_a <- id_plot_legis_dyn(unemp1_fit,include="AMASH, Justin",
                         use_ci=T,plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Spline of 2nd Degree") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))
  
  saveRDS(un1_a, "data/un1_a.rds")

un2_a <- id_plot_legis_dyn(unemp2_fit,use_ci=T,include="AMASH, Justin",plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Spline of 3rd Degree") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))
 
  saveRDS(un2_a, "data/un2_a.rds")

un3_a <- id_plot_legis_dyn(unemp3_fit,use_ci=T,include="AMASH, Justin",plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Spline of 4th Degree") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))
  
  saveRDS(un3_a, "data/un3_a.rds")

gp_a <- id_plot_legis_dyn(unemp_gp_fit,use_ci=T,include="AMASH, Justin",plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Gaussian Process") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))

  saveRDS(gp_a, "data/gp_a.rds")

ar_a <- id_plot_legis_dyn(unemp_ar_fit,use_ci=T,include="AMASH, Justin",plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("AR(1) Process") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))

  saveRDS(ar_a, "data/ar_a.rds")

rw_a <- id_plot_legis_dyn(unemp_rw_fit,use_ci=T,include="AMASH, Justin",plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Random Walk") +
  scale_x_date(guide = guide_axis(n.dodge = 2)) + labs(y="")

  saveRDS(rw_a, "data/rw_a.rds")
  
  un1_am <- id_plot_legis_dyn(unemp1_fitm,include="AMASH, Justin",
                         use_ci=T,plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Spline of 2nd Degree") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))
  
  saveRDS(un1_am, "data/un1_am.rds")

un2_am <- id_plot_legis_dyn(unemp2_fitm,use_ci=T,include="AMASH, Justin",plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Spline of 3rd Degree") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))
 
  saveRDS(un2_am, "data/un2_am.rds")

un3_am <- id_plot_legis_dyn(unemp3_fitm,use_ci=T,include="AMASH, Justin",plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Spline of 4th Degree") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))
  
  saveRDS(un3_am, "data/un3_am.rds")

gp_am <- id_plot_legis_dyn(unemp_gp_fitm,use_ci=T,include="AMASH, Justin",plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Gaussian Process") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))

  saveRDS(gp_am, "data/gp_am.rds")

ar_am <- id_plot_legis_dyn(unemp_ar_fitm,use_ci=T,include="AMASH, Justin",plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("AR(1) Process") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))

  saveRDS(ar_am, "data/ar_am.rds")

rw_am <- id_plot_legis_dyn(unemp_rw_fitm,use_ci=T,include="AMASH, Justin",plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Random Walk") +
  scale_x_date(guide = guide_axis(n.dodge = 2)) + labs(y="")

  saveRDS(rw_am, "data/rw_am.rds")
  
# Now do Tim Walz  
  
un1_tw <- id_plot_legis_dyn(unemp1_fit,include="WALZ, Tim",
                         use_ci=T,plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Spline of 2nd Degree") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))
  
  saveRDS(un1_tw, "data/un1_tw.rds")

un2_tw <- id_plot_legis_dyn(unemp2_fit,use_ci=T,include="WALZ, Tim",plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Spline of 3rd Degree") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))
 
  saveRDS(un2_tw, "data/un2_tw.rds")

un3_tw <- id_plot_legis_dyn(unemp3_fit,use_ci=T,include="WALZ, Tim",plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Spline of 4th Degree") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))
  
  saveRDS(un3_tw, "data/un3_tw.rds")

gp_tw <- id_plot_legis_dyn(unemp_gp_fit,use_ci=T,include="WALZ, Tim",plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Gaussian Process") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))

  saveRDS(gp_tw, "data/gp_tw.rds")

ar_tw <- id_plot_legis_dyn(unemp_ar_fit,use_ci=T,include="WALZ, Tim",plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("AR(1) Process") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))

  saveRDS(ar_tw, "data/ar_tw.rds")

rw_tw <- id_plot_legis_dyn(unemp_rw_fit,use_ci=T,include="WALZ, Tim",plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Random Walk") +
  scale_x_date(guide = guide_axis(n.dodge = 2)) + labs(y="")

  saveRDS(rw_tw, "data/rw_tw.rds")
  
# Tim Walz with missing data
  
un1_twm <- id_plot_legis_dyn(unemp1_fitm,include="WALZ, Tim",
                         use_ci=T,plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Spline of 2nd Degree") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))
  
  saveRDS(un1_twm, "data/un1_twm.rds")

un2_twm <- id_plot_legis_dyn(unemp2_fitm,use_ci=T,include="WALZ, Tim",plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Spline of 3rd Degree") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))
 
  saveRDS(un2_twm, "data/un2_twm.rds")

un3_twm <- id_plot_legis_dyn(unemp3_fitm,use_ci=T,include="WALZ, Tim",plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Spline of 4th Degree") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))
  
  saveRDS(un3_twm, "data/un3_twm.rds")

gp_twm <- id_plot_legis_dyn(unemp_gp_fitm,use_ci=T,include="WALZ, Tim",plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Gaussian Process") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))

  saveRDS(gp_twm, "data/gp_twm.rds")

ar_twm <- id_plot_legis_dyn(unemp_ar_fitm,use_ci=T,include="WALZ, Tim",plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("AR(1) Process") + labs(y="") +
  scale_x_date(guide = guide_axis(n.dodge = 2))

  saveRDS(ar_twm, "data/ar_twm.rds")

rw_twm <- id_plot_legis_dyn(unemp_rw_fitm,use_ci=T,include="WALZ, Tim",plot_text = F) + scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + ggtitle("Random Walk") +
  scale_x_date(guide = guide_axis(n.dodge = 2)) + labs(y="")

  saveRDS(rw_twm, "data/rw_twm.rds")
  
# need to export covariates, anything else we might need
  
s1_cov <- unemp1_fit@stan_samples$draws("legis_x") %>% 
  summarize_draws %>% 
  mutate(Model="Spline 2nd Degree",Missingness="Unmodeled")
s2_cov <- unemp2_fit@stan_samples$draws("legis_x") %>% 
  summarize_draws %>% 
  mutate(Model="Spline 3rd Degree",Missingness="Unmodeled")
s3_cov <- unemp3_fit@stan_samples$draws("legis_x") %>% 
  summarize_draws %>% 
  mutate(Model="Spline 4th Degree",Missingness="Unmodeled")
sgp_cov <- unemp_gp_fit@stan_samples$draws("legis_x") %>% 
  summarize_draws %>% 
  mutate(Model="Gaussian Process",Missingness="Unmodeled")
sar_cov <- unemp_ar_fit@stan_samples$draws("legis_x") %>% 
  summarize_draws %>% 
  mutate(Model="AR(1)",Missingness="Unmodeled")
srw_cov <- unemp_rw_fit@stan_samples$draws("legis_x") %>% 
  summarize_draws %>% 
  mutate(Model="Random Walk",Missingness="Unmodeled")

s1_covm <- unemp1_fitm@stan_samples$draws("legis_x") %>% 
  summarize_draws %>% 
  mutate(Model="Spline 2nd Degree",Missingness="Modeled")
s2_covm <- unemp2_fitm@stan_samples$draws("legis_x") %>% 
  summarize_draws %>% 
  mutate(Model="Spline 3rd Degree",Missingness="Modeled")
s3_covm <- unemp3_fitm@stan_samples$draws("legis_x") %>% 
  summarize_draws %>% 
  mutate(Model="Spline 4th Degree",Missingness="Modeled")
sgp_covm <- unemp_gp_fitm@stan_samples$draws("legis_x") %>% 
  summarize_draws %>% 
  mutate(Model="Gaussian Process",Missingness="Modeled")
sar_covm <- unemp_ar_fitm@stan_samples$draws("legis_x") %>% 
  summarize_draws %>% 
  mutate(Model="AR(1)",Missingness="Modeled")
srw_covm <- unemp_rw_fitm@stan_samples$draws("legis_x") %>% 
  summarize_draws %>% 
  mutate(Model="Random Walk",Missingness="Modeled")

cov_combined <- bind_rows(s1_cov, s2_cov, s3_cov,
                          sgp_cov, sar_cov, srw_cov,
                          s1_covm, s2_covm, s3_covm,
                          sgp_covm, sar_covm, srw_covm)

saveRDS(cov_combined, "data/cov_combined.rds")

rm(list=ls()[grepl(x=ls(),pattern = "unemp")])
  
  
}

```

The expected posterior distributions of the ideal point scores over time for each model are shown in @fig-alldist when employing only observed votes (i.e., ignoring absences and abstentions). As can be seen, the different methods have differing interpretations of ideal point trajectories, although all models in general show relative stability over the monthly periods of the 115th Congress. The main difference is that the more complex models, including the AR(1), random walk, and GP, show much higher levels of over-time variation for some legislators. Based on the short time period of the 115th Congress, and the fact that we are attempting to study legislator ideology, we should be skeptical of these rapid trajectories as reflecting either noise in the data or other aspects of legislative behavior (such as the scheduling of votes) that are polluting our estimate of the latent trait. The spline-based models, by contrast, show change over time for some legislators but change is relatively smooth, i.e., ideology does not change rapidly from to rollcall to rollcall as we would expect.

```{r dists}
#| label: fig-alldist
#| fig-cap: "Comparison of Time-varying Ideal Point Methods for the 115th Congress with Observed Votes"
#| message: false
#| warning: false

  un1 <- readRDS("data/un1.rds")
  un2 <- readRDS("data/un2.rds")
  un3 <- readRDS("data/un3.rds")
  rw <- readRDS("data/rw.rds")
  ar <- readRDS("data/ar.rds")
  gp <- readRDS("data/gp.rds")

un1 + un2 + un3 + ar + gp + rw + plot_layout(nrow=2,
                                             guides="collect") + 
  plot_annotation(caption="Posterior average values shown for 115th Congress legislator ideal points.\nItem discrimination for Republican party-line votes was constrained to be positive.") & theme(legend.position = "top",
                                                                                                                                                                                                    plot.title=element_text(size=9))


```

```{r distsamash1}
#| label: fig-amashdist
#| fig-cap: "Comparison of Justin Amash Time-varying Ideal Points for the 115th Congress with Observed Votes"
#| message: false
#| warning: false

  un1_a <- readRDS("data/un1_a.rds")
  un2_a <- readRDS("data/un2_a.rds")
  un3_a <- readRDS("data/un3_a.rds")
  gp_a <- readRDS("data/gp_a.rds")
  rw_a <- readRDS("data/rw_a.rds")
  ar_a <- readRDS("data/ar_a.rds")

un1_a + un2_a + un3_a + ar_a + gp_a + rw_a + plot_layout(nrow=2,
                                             guides="collect") + 
  plot_annotation(caption="Posterior average values shown for 115th Congress legislator ideal points.\nItem discrimination for Republican party-line votes was constrained to be positive.") & theme(legend.position = "none",
                                                                                                                                                                                                    plot.title=element_text(size=9))

```

Even with the relative stability in @fig-alldist, there are some legislators whose ideal points do change over time. In @fig-amashdist, I plot the monthly ideal point trajectory for Justin Amash, a Republican Congressman who famously became embroiled in disputes with then-President Donald Trump over foreign policy. As @fig-amashdist shows, Justin Amash shows strong movement in a positive or conservative direction in the middle of the Congressional session. This seems to correspond with his increasing disputes with the Trump White House over foreign policy in particular. Amash's criticism escalated after Trump had a closed door meeting with Russian president Vladimir Putin on July 16, 2016.[^5] Per the plots in @fig-amashdist, this event coincided with noticeable movement in a conservative direction. All of the models show this movement, while the more complex time series like GPs show a more sudden jump. The splines, despite being much simpler, are able to capture the approximate same trajectory while using fewer parameters.

[^5]: For more information, see \url{https://thehill.com/homenews/house/397787-gop-lawmaker-trump-went-out-of-his-way-to-appear-subordinate-at-putin-press/}.

We can next turn to estimation of legislator ideal points for the 115th Congress when incorporating strategic abstention. To provide context of abstention in the U.S. House, Figures 3 and 4 in the SI provide descriptive information about absence rates in the U.S. House from 1990 to 2018 measured for each rollcall vote. The figures reveal that while absence rates are stable over time, they tend to increase towards the end of the legislative session. This within-session variation in absence rates implies that a time-varying model that incorporates this missing data mechanism—i.e. to discover whether legislators tend to show up for rollcall votes that are more or less polarized—could likewise capture new insights into the study of strategic legislative absence [@cloléry2023; @battaglini2023; @rothenberg2000].

@fig-alldist2 shows that there is justification for non-ignorable missing data (i.e. strategic abstention) in the 115th Congress. The spline models show interesting heterogeneity at the end of the session. In general, the more complex models show even more over-time variation, though the movements are so large as to be implausible and are another reason that simpler splines should be preferred given the nature of the theoretical construct. There is also some evidence of increasing overlap in the distributions between parties at the end of the session which is not apparent in the observed data distribution in @fig-alldist. This could be evidence that at the end of the session there is increased opportunity for deal-making.

```{r dists2}
#| label: fig-alldist2
#| fig-cap: "Comparison of Time-varying Ideal Point Methods for the 115th Congress with Missing Votes"
#| message: false
#| warning: false

  un1m <- readRDS("data/un1m.rds")
  un2m <- readRDS("data/un2m.rds")
  un3m <- readRDS("data/un3m.rds")
  rwm <- readRDS("data/rwm.rds")
  arm <- readRDS("data/arm.rds")
  gpm <- readRDS("data/gpm.rds")

un1m + un2m + un3m + arm + gpm + rwm + plot_layout(nrow=2,
                                             guides="collect") + 
  plot_annotation(caption="Posterior average values shown for 115th Congress legislator ideal points.\nItem discrimination for Republican party-line votes was constrained to be positive.") & theme(legend.position = "top",
                                                                                                                                                                                                    plot.title=element_text(size=9))

```

In SI Section 7 I also plot the monthly ideal points for then-Representative Tim Walz with only observed data and in @fig-walzdist2 the ideal points for Walz, who in 2018 announced his run to be governor of Minnesota. While the observed data models reveal little change after his gubernatorial announcement, the missing data models show show his ideal point moving in a clear negative or liberal direction after he started to run for governor. Once it became more costly for Walz to show up to vote, his pattern of voting became more similar to those who were on the left wing of the party. This behavior could correspond to Walz's need to appear more liberal in order to win the Democratic primary in the gubernatorial race, which occurred on August 15, 2018.

```{r distswalz2}
#| label: fig-walzdist2
#| fig-cap: "Comparison of Tim Walz Time-varying Ideal Points for the 115th Congress with Missing Votes"
#| message: false
#| warning: false

  un1_twm <- readRDS("data/un1_twm.rds")
  un2_twm <- readRDS("data/un2_twm.rds")
  un3_twm <- readRDS("data/un3_twm.rds")
  gp_twm <- readRDS("data/gp_twm.rds")
  rw_twm <- readRDS("data/rw_twm.rds")
  ar_twm <- readRDS("data/ar_twm.rds")

un1_twm + un2_twm + un3_twm + ar_twm + gp_twm + rw_twm + plot_layout(nrow=2,
                                             guides="collect") + 
  plot_annotation(caption="Posterior average values shown for 115th Congress legislator ideal points.\nItem discrimination for Republican party-line votes was constrained to be positive.") & theme(legend.position = "none",
                                                                                                                                                                                                    plot.title=element_text(size=9))

```

```{r legisx}
#| label: fig-coef
#| tbl-cap: Coefficients of GOP and District Unemployment Interaction by Model Type ($\phi$)
#| cache: true
#| eval: false

library(tinytable)
library(dplyr)

readRDS("data/cov_combined.rds") %>% 
  bind_rows %>% 
  mutate(variable=recode(variable,
                         `legis_x[1]`="Unemployment",
                         `legis_x[2]`="GOP",
                         `legis_x[3]`="UnemploymentXGOP")) %>% 
  filter(variable %in% c("Unemployment",
                         "UnemploymentXGOP")) %>% 
  select(Model,Variable="variable",
         Missingness,
         `Posterior Median`="median",
         `5% Quantile`="q5",
         `95% Quantile`="q95") %>% 
  ggplot(aes(y=`Posterior Median`,x=Model)) +
  geom_pointrange(aes(ymin=`5% Quantile`, 
                      ymax=`95% Quantile`,
                      linetype=Missingness),
                  position=position_dodge(0.5),colour="black") +
  facet_wrap(~Variable,nrow=2) +
  geom_hline(yintercept=0,linetype=3,colour="black") +
  ggthemes::theme_clean() +
  labs(caption=stringr::str_wrap("Plot shows hierarchical covariates that predict ideal points. Estimates are organized by the type of latent time process and by whether missing data was incorporated into the model. Coefficient estimates are available as a table in the supplementary information.",width=55)) +
  coord_flip()
  # arrange(Variable,Model) %>% 
  # filter(Variable %in% c("Unemployment",
  #                        "GOP",
  #                        "UnemploymentXGOP")) %>% 
  # tt(digits=3,
  #    notes="Plot shows hierarchical legislator-levle covariates for different ideal point models, including time functions and whether or not the model accounted for missingness.") %>% 
  # style_tt(i=1:36,
  #          fontsize=.7)

```

```{r loadmarg}
#| include: false
#| eval: false

by_party <- readRDS("data/by_party.rds")

```

```{r calcmargeffs}
#| label: fig-allplot
#| fig-cap: "Item-level Ideal Point Marginal Effects of Monthly District Unemployment on Legislator Ideal Points"
#| out-width: 100%
#| fig-height: 6
#| eval: false

c1 <-   by_party %>% 
    mutate(group_id=factor(group_id,levels=c("D","R"),
                           labels=c("Democrats", "Republicans"))) %>% 
    #filter(!(item_orig %in% c("115_1050","115_588"))) %>% 
    ggplot(aes(y=mean_est,
               x=reorder(item_id,mean_est))) +
    geom_linerange(aes(ymin=low_est,
                       ymax=high_est,
                       colour=`median`)) +
    facet_wrap(~group_id) +
    scale_colour_viridis_c(name="Discrimination") +
    coord_flip() +
    labs(y="Marginal Change in Probability of Voting",
         x="Rollcalls",
         caption="Marginal effect of unemployment on voting on a specific rollcall in the 115th Congress.\nEach rollcall uncertainty interval is colored by the average posterior discrimination of that rollcall.\nMore positive rollcalls tend to be more conservative votes,\nwhile more negative rollcalls tend to be more liberal votes.") +
    geom_hline(yintercept=0,linetype=2) +
    ggthemes::theme_tufte() +
    theme(axis.text.y=element_blank(),
          axis.ticks.y=element_blank(),
          axis.line = element_blank(),
          panel.grid=element_blank(),
          legend.position = "bottom")

ggsave("idealpt_marg_eff.png",plot=c1,width=6,height=4)

c2 <- c1 + ggdark::dark_mode() + 
  theme(axis.text.y=element_blank(),
          axis.ticks.y=element_blank(),
          axis.line = element_blank()) +
  labs(y="Marginal Change in Probability of Voting",
         x="Rollcalls",
         caption="Marginal effect of unemployment on voting on a specific rollcall in the 115th Congress.\nEach rollcall uncertainty interval\nis colored by the average posterior discrimination of that rollcall.\nMore positive rollcalls tend to be more conservative votes,\nwhile more negative rollcalls tend to be more liberal votes.")

ggsave("idealpt_marg_eff_dark.png",plot=c2,width=6,height=4)
  
print(c1)
```

```{r calcmargeffsp,include=FALSE, eval=FALSE}
#| label: fig-allplotp
#| eval: false
#| fig-cap: "Comparison of Item-level Ideal Point Marginal Effects of Unemployment on Legislator Ideal Points and Party Moderation"

by_party %>% 
    ungroup %>% 
    mutate(group_id=factor(group_id,levels=c("D","R"),
                           labels=c("Democrats", "Republicans")),
           item_rank=rank(mean_est)) %>% 
    filter(!(item_orig %in% c("115_1050","115_588"))) %>% 
    ggplot(aes(y=mean_est,
               x=item_rank)) +
    geom_ribbon(aes(ymin=low_est,
                    ymax=high_est,
                    fill=group_id),alpha=0.5) +
    #facet_wrap(~group_id) +
    ggthemes::theme_tufte() + 
    scale_fill_manual(values=c(Republicans="red",
                              Democrats="blue"),name="") +
    coord_flip() +
    labs(y="Marginal Change in Probability of Voting",
         x="Rollcalls") +
    geom_hline(yintercept=0,linetype=2) +
    ggdark::dark_mode() +
    theme(axis.text.y=element_blank(),
          axis.ticks.y=element_blank()) +
    ggtitle("Marginal Effect of District Monthly Unemployment on Rollcall Votes in 115th Congress",
            subtitle="Marginal Effect of Unemployment Mediated by Legislator Ideal Point and Bill Discrimination")

```

## Full Trajectories of U.S. House from 1990 to 2018

In this section, I apply the spline-based models to all House sessions from 1990 to 2018. The complete distributions for all legislators for both observed-data and missing-data models are shown in @fig-fulldist. These plots show a similar stability of over-time trends comparable to the much shorter time series in SI Figure 10, which indicates that Congressional ideal points do not change very rapidly. There does not appear to be significantly more or less movement as the degree of the spline increases. There is some additional movement in the missing-data models, though not of a very extreme nature.

![Monthly Ideal Points for U.S. House Legislators from 1990 to 2018](full_dist.pdf){#fig-fulldist}

In @fig-sumbigdist I show the party-level ideal points for both Democrats and Republicans over the same time period. To derive these scores, the individual-level ideal point trajectories are first aggregated to the group level, and then averages are taken of all posterior draws. This method preserves uncertainty in the posterior expectation, although the uncertainty intervals are not shown to make the differences between models more apparent.

```{r}
#| fig-cap: Monthly Party-level Ideal Points for the U.S. House
#| label: fig-sumbigdist
#| fig-height: 4

# load all these distributions, plot them 

all_group_big <- lapply(list.files("data/","idealpts",full.names = T), 
                        readRDS) %>% 
  bind_rows(.id="model") %>% 
  mutate(Spline=case_match(as.character(model),
                          c("1","2")~"2°",
                          c("3","4")~"3°",
                          c("5","6")~"4°"),
         Type=case_match(model,
                          c("2","4","6")~"With Abstentions",
                          c("1","3","5")~"Only Observed\nVotes"))

# make labels 

annotations <- tibble(x=rep(c(ymd("1993-01-01"),ymd("1993-01-01")),2),
           y=rep(c(-7,7),2),
           label=rep(c("Democrats","Republicans"),2),
           Type=rep(unique(all_group_big$Type),each=2))

all_group_big %>% 
  ggplot(aes(y=mean_est,x=Time_Point)) +
  geom_line(aes(linetype=Spline,colour=Group)) +
  facet_wrap(~Type,nrow=2) + 
  ggthemes::theme_clean() +
  scale_color_manual(values=c(R="red",
                              D="blue"),na.translate=F,name="") + 
  guides(color="none",
         linetype=guide_legend(override.aes = list(color = "black"))) + 
  scale_y_continuous(labels=c("More\nLiberal","-5","0","5","More\nConservative"),breaks=c(-10,-5,0,5,10)) +
  geom_text(data=annotations,aes(label=label,x=x,y=y),colour="black") +
  labs(y="Ideal Point",x="",
       caption=stringr::str_wrap("Plot shows party-average monthly ideal points for the U.S. House from 1990 to 2018 that are aggregated from individual legislator trajectories.")) 

#  Ideal point is derived from a Bayesian IRT model fit with idealstan. Missing-data models account for strategic abstention while observed models only use recorded votes. Each line is an estimate from a different type of spline model in terms of degrees with knots for each presidential administration. Discrete jumps in the time series show periods when new legislators entered.



```

# Discussion

What is clear is that @fig-sumbigdist departs noticeably from the DW-NOMINATE scores that generally show a GOP that is moving away from Democrats since 1990. Instead, it would appear that Democrats were moving away from Republicans after 1990. This polarization reached a peak in 2012, and then moderated until the end of 2018. The missing-data models that incorporate strategic abstention show much greater moderation, which could be due to either increasing strategic abstention or the artificial end of the time series in 2018 (or a combination of both factors). However, despite these differences, both the observed data and missing-data models show a similar trajectory. Furthermore, there are very minor differences between the different spline models in terms of degrees and the flexibility of ideal points over time.

It is worth considering the differences between these scores and those of DW-NOMINATE. DW-NOMINATE party-level ideal points are based on legislator ideal points that vary linearly and are estimated once for each Congressional session. Furthermore, these ideal point trajectories are scaled to fit within a unit hypersphere for identification purposes [@poole1997]. As such, there is considerable stretching of the ideal points to fit within these constraints. By contrast, in the `idealstan` model, constraints are placed on the item (vote) parameters while the ideal points are allowed to float with a weakly informative prior. As such, there is reason to believe that the trajectories in @fig-sumbigdist are closer to the actual political trajectories of the parties.

There are obviously many interesting hypotheses that can follow from these new party-level and individual-level measures for the U.S. House, and I leave these questions for scholars to pursue as these estimates will be available for further study.

```{=html}
<!-- # Conclusion

In this paper, I presented a generalization of the increasingly popular ideal point model in the Bayesian IRT framework. I extended the model with new modes of missing-data, time-varying processes, and mixed outcomes. The main contribution of this paper is an analytical tool that can extend the domain and applicability of ideal point models across the discipline as all of these models are available in a single R package, `idealstan`. Crucially, all of these methods build on each other so that missing data can be incorporated in time-varying models or in joint distributions and vice versa.

The package `idealstan` is designed to automate the sometimes arduous process of preparing data for ideal point modeling, including identifying parameters. Furthermore, the use of a single R package for all models enables researchers to compare different models on the same data with ease. Parallelization of standard Bayesian inference is also incorporated in the package to enable estimation of the ever larger data sets available to political scientists. -->
```

# References